---
title: "Timeseries"
output:
  pdf_document: default
  html_document: default
date: "2023-08-07"
---

```{r setup, results='hide'}
knitr::opts_chunk$set(echo = TRUE)
```

## Time Series

```{r, results='hide',message=FALSE}
#libraries
library(readxl)
library(lubridate)
library(zoo)
library(xts)
library(forecast)
library(dplyr)
```

```{r}
dir_name <- "Data Sources"
file_name <- "NY-JFKairport-temperatures.csv"
file_path <- paste(getwd(),dir_name,file_name,sep="/")
```

```{r}
df_monthly <- read.csv(file_path)
```

#### Including Plots

You can also embed plots, for example:

```{r}
head(df_monthly)
```
```{r}
df_monthly$DATE <- as.Date(as.yearmon(df_monthly$DATE))
df_monthly$Year <- floor_date(df_monthly$Date)
```

```{r pressure, results='hide'}
plot(x = df_monthly$DATE, y = df_monthly$TAVG, type = "l")
```

### Issue with Time and How to Solve it
1. Take consideration of Time Zone
2. Daylight Savings Time

How?

Use UTC or GMT Time
```{r}
getwd()
```

```{r}
dir_name <- "Data Sources"
file_name_e <- "Raleigh-Durham_North_Carolina_area_electricity_demand.csv"
file_path_e <- paste(getwd(),dir_name,file_name_e,sep="/")
elec <- read.csv(file_path_e)
```

```{r}
elec$LocalDatetime <- as.POSIXct(elec$Datetime, tz = "America/New_York")
elec$UTC <- as_datetime(elec$Datetime)
```

```{r}
zoo_monthly <- zoo(df_monthly$TAVG, df_monthly$DATE)
xts_monthly <- xts(df_monthly$TAVG, df_monthly$DATE)
```

```{r}
colnames(xts_monthly) <- c('Temperature')
#convert xts to zoo
#as.zoo(xts_monthly)
```

```{r}
plot(xts_monthly, main = "Average Monthly Temperature", xlab=  "Month-Year",ylab = "Temperature(degrees Fahrenheit)", ylim = c(20,90))
```

Notice hotter temp in hotter months, lower temp in colder months

```{r}
#last 10 years
plot(tail(zoo_monthly,120), main = "Average Monthly Temperature", xlab=  "Month-Year",ylab = "Temperature(degrees Fahrenheit)", ylim = c(20,90))
```
```{r}
xts_electricity <- xts(elec$Demand, elec$LocalDatetime)
#in xts function first argument goes to the column and second goes to row
colnames(xts_electricity) <- c('Demand')
```

```{r}
plot(tail(xts_electricity,24*7)) #how it oscillates for each week
```

```{r}
zoo_elec <- zoo(elec$Demand, elec$UTC)
plot(zoo_elec)
```

xts and zoo elec won't line up because xts used local time and zoo used UTC time!!

```{r}
df_monthly_again <- as.data.frame(xts_monthly)
df_monthly_again$Date <- rownames(df_monthly_again)
```


#### Look at Subset of the data

Filtering data that we want to look at
```{r}
df_monthly[df_monthly$TAVG == 76.9,]
```

#### Window function
works on xts and zoo
```{r}
xts_monthly_2010s <- window(xts_monthly,start = "2010-01-01", end = "2019-12-31")
```

```{r}
plot(window(xts_electricity, start = "2019-03-07", end="2019-03-14"))
```

Notice that there isn't any data

#### how to fill in the empty value
in excel it doesn't consider empty values
```{r}
window(zoo_monthly,start = "1948-07-01", end = "1948-12-01")
```
```{r}
mean(window(zoo_monthly,start = "1948-07-01", end = "1948-12-01"))
```
```{r}
mean(window(zoo_monthly,start = "1948-08-01", end = "1948-12-01"))
```
```{r}
summary(xts_monthly)
```
Notice there are 2 NA's only
```{r}
xts_monthly <- na.omit(xts_monthly)
```

#### But how do you fill in the empty ones?
```{r}
plot(na.locf(window(zoo_elec, start = "2019-03-07", end="2019-03-14")))
```
This carries last observation forward until the new data value and fill in the NA with it
```{r}
zoo_update <-plot(window(xts_electricity, start = "2019-03-07", end="2019-03-14"))
```

#### Pivot table
```{r}
apply.daily(xts_electricity,sum)
```
```{r}
apply.weekly(xts_electricity,sum)
```
```{r,results='hide'}
apply.monthly(xts_electricity,sum)
```
```{r}
apply.quarterly(xts_monthly,mean)
```

```{r}
xts_yearly <- apply.yearly(xts_monthly,start = "1949-01-01", end = "2021-12-31",mean)
```

#### Line chart breakdown
```{r}
plot(decompose(ts(xts_monthly,frequency = 12)),ylim=c(30,100))
```

R doesn't know frequency --> convert time series with ts function

1. Original Time series
2. Trend
fluctuate 50 ~58 --> smaller range
narrower b/c it removes seasonality from the range
3. Seasonal
Look at the diff between hot and cold weather
4. Random
represent abnormalities in the data(look at the spike)


```{r}
plot(decompose(ts(window(xts_monthly,start = "2010-01-01", end = "2019-12-31"),frequency = 12)))
```

## ARIMA Model
Autoregressive integrated moving average
=arima(<time series object>, order = c(p,d,q))
p,d,q vector

ARIMA models: fitted

```{r}
results_yearly <-arima(xts_yearly,order=c(0,0,0))
```

```{r}
mean(xts_yearly)
```

```{r}
xts_yearly_fitted <- merge(xts_yearly, xts(fitted(results_yearly),index(xts_yearly)))
colnames(xts_yearly_fitted) <- c('Temperature','Fitted')
```

```{r}
plot(xts_yearly_fitted,col=c('black','orange'))
```
#### calculate std dev
```{r}
sd(xts_yearly)
```
```{r}
predict(results_yearly,10)
```
```{r}
plot(forecast(results_yearly,10), xaxt = 'n', xlab="Year",ylab="Temperature (degrees F)")
axis(1, at = c(2,12,22,32,42,52,62,72), labels=c(1950,1960,1970,1980,1990,2000,2010,2020))
#xaxt='n' removes tickmarks
#reference x axis = axis(1,)
```
Forecasted future temp as Blue line
80% CI blue box, 90% CI grey box


```{r}


```



Aggregating data -> github.com/helenrmwall/Time-Series-Models
Use hourly electricity demand for San Diego stored on GitHub and in exercise files
Determine which week over an entire time period has highest total electricity demand
set tz = "America/Los_Angeles"

```{r}
file_name_sd <- "San Diego electricity demand.csv"
file_path_sd <- paste(getwd(),dir_name,file_name_sd,sep="/")
df_sandiego <- read.csv(file_path_sd)
```
```{r}
df_sandiego$LocalDateTime <- as.POSIXct(df_sandiego$Datetime,tz="America/Los_Angeles")
```
```{r}
xts_sandiego <- xts(df_sandiego$Demand,df_sandiego$LocalDateTime)
colnames(xts_sandiego) <- c('demand')
```
```{r}
plot(apply.weekly(xts_sandiego,sum))
```

```{r}
df_sandiego_weekly <- as.data.frame(apply.weekly(xts_sandiego,sum))
df_sandiego_weekly$Date <- as.Date(rownames(df_sandiego_weekly))
df_sandiego_weekly[df_sandiego_weekly$demand == max(df_sandiego_weekly$demand),]
```

#### Linear Regression
```{r}
m <- lm(coredata(xts_yearly) ~ index(xts_yearly))
xts_yearly_lm <- merge(xts_yearly,xts(predict(m, newdata = xts_yearly, response="type"),index(xts_yearly)))
colnames(xts_yearly_lm) <- c('Temperature','Fitted')
```

```{r}
plot(xts_yearly_lm, col=c('black','orange'))
```

#### Residuals
```{r}
residuals <- xts(resid(m),index(xts_yearly))
plot(zoo(residuals))
abline(h=0,col='orange',lwd=3)
```

#### Lag
Delay or move a time series by n periods of time
Line up separate time data observatiosn
Bootstrap dat within single time series

```{r, results='hide'}
lead(xts_yearly,1)
#lag(xts_yearly,1)
```

```{r}
xts_monthly_lagged <- xts_monthly
for (i in 1:12) {
  xts_monthly_lagged <- merge(xts_monthly_lagged,lag(xts_monthly,i))
}
```


#### Autoregression
Autoregression coefficient of 1 = linear fit of current point and the point right before it
= 2 points & 1 line

```{r}
results_yearly <-arima(xts_yearly,order=c(2,0,0))
xts_yearly_fitted <- merge(xts_yearly, xts(fitted(results_yearly),index(xts_yearly)))
colnames(xts_yearly_fitted) <- c('Temperature','Fitted')
plot(xts_yearly_fitted,col=c('black','orange'))
```

#### Moving Averages, Rolling Averages
Even out fluctuations within time seires models
moving average of period of 2 --> use from one previous to one future (total of 3)

```{r}
xts_yearly_rolling <- merge(xts_yearly,rollmean(xts_yearly,2,align='left'))
plot(xts_yearly_rolling,col=c('black','orange'))
```

```{r}
plot(xts_sandiego)
```

```{r}
m_sandiego <- lm(coredata(xts_sandiego)~index(xts_sandiego))
m_sandiego
```
```{r}
xts_sandiego_lm <- merge(xts(predict(m_sandiego,newdata=xts_sandiego,response='type'),index(xts_sandiego)),xts_sandiego)
```

```{r}
colnames(xts_sandiego_lm) <- c('Fitted','Demand')
plot(xts_sandiego_lm,col=c('black','grey'))
```

#### Autocorrelation
```{r}
plot(xts_monthly)
pairs(as.data.frame(na.omit(xts_monthly_lagged)),xaxt='n',yaxt='n',pch=20)
```
```{r}
#cor(na.omit(xts_monthly_lagged))
acf(xts_monthly,plot=FALSE,lag.max = 12)
```
This compares value of current to the subsequent month. For example, it compares 1st month to 2nd month, 2nd to 3rd, ...

#### Partial Autocorrelation
Steps between the lags of the model

```{r}
pacf(xts_monthly)
```
```{r}
pacf(xts_monthly,plot=FALSE,lag.max=12)
```
#### Stationarities
```{r}
plot(xts_yearly)
```
```{r}
plot(diff(zoo(xts_yearly)))
abline(h=0,col='orange',lwd=3)
```

```{r}
results_yearly <-arima(xts_yearly,order=c(0,1,0))
xts_yearly_fitted <- merge(xts_yearly, xts(fitted(results_yearly),index(xts_yearly)))
colnames(xts_yearly_fitted) <- c('Temperature','Fitted')
plot(xts_yearly_fitted,col=c('black','orange'))
```
```{r}
plot(forecast(results_yearly,10), xaxt = 'n', xlab="Year",ylab="Temperature (degrees F)")
axis(1, at = c(2,12,22,32,42,52,62,72), labels=c(1950,1960,1970,1980,1990,2000,2010,2020))
```

```{r}
pacf(xts_yearly)
```
1st 0.4, 4th 0.3
```{r}
acf(xts_yearly)
```
first 2 pair strong coefficient
```{r}
results_yearly <-auto.arima(xts_yearly)
results_yearly
```
p = 0 d = 1 q = 2

```{r}
confint(results_yearly)
```

```{r}
xts_yearly_fitted <- merge(xts_yearly, xts(fitted(results_yearly),index(xts_yearly)))
colnames(xts_yearly_fitted) <- c('Temperature','Fitted')
plot(xts_yearly_fitted,col=c('black','orange'))
```
```{r}
auto.arima(xts_monthly,D=12, max.p=24, max.q=24)
```




